Measuring causal influence with\\ back-to-back regression

### 2019-05-23 paper/paper.tex

% TODO: Talk about cause-effect methods, trace method, half-sibling
% TODO: Cite Roweis component analysis methods paper
% TODO: review abstract
% TODO: review conclusion

@@

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx}
\usepackage[ruled, linesnumbered]{algorithm2e}

\usepackage[hidelinks]{hyperref}
\usepackage[]{neurips_2019}
\DeclareMathOperator{\Tr}{Tr}

@@

If using regularized least-squares, arguably the most commonly employed linear regression technique [CITE], the solution of B2BR has a closed form solution:
\begin{align}
    \hat G &= (Y^\top Y + \Lambda_Y)^{-1} Y^\top X,\label{eq:solG}\\
    \hat H &=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G,\label{eq:solH}
\end{align}
%
where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices of regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split [CITE BOOTSTRAP].
%
After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.

@@

Recall the rank factorisation property of the pseudo-inverse [CITE]: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
%
Using this property on the minimal norm solution $\hat G$ yields
\begin{equation}
\begin{aligned}
(XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
&=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
&=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
\end{aligned}
\end{equation}

@@

In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero [CITE].
%
However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
%
Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
%
More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan [CITE] criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{E}$, set to one all the diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the diagonal entries.

@@

Back-to-back regression is closely related to Canonical Correlation Analysis [CITE].
%
While CCA computes the eigenvector decomposition of \eqref{eq:solH}, while we compute its diagonal.
%
If \eqref{eq:solH} is diagonal, as in our model \eqref{eq:model}, the variables in $X$ are the eigenvectors.
%
If all the associated eigenvalues are either $0$ or $1$, then all the variables are either part of the $Y$ subspace, or orthogonal to it, respectively.
%
This is because these eigenvalues correspond to the squared cosine of the angle between the corresponding variable in $X$ and the $Y$ subspace.
%
Thus, back-to-back regression is equivalent to CCA under the constraint that the subspaces spanned by $X$ and $Y$ are perpendicular, $X$ is spanned by vectors of $Y$ and vectors orthogonal to $Y$, and that the parallel and orthogonal components of $X$ are spanned by disjoint sets of features of $X$.

@@

\bibliographystyle{abbrvnat}
\bibliography{paper}

>>>>

% TODO: Talk about cause-effect methods, trace method, half-sibling
% TODO: Cite Roweis component analysis methods paper
% TODO: read quantifying causal influences
% TODO: review abstract
% TODO: review conclusion

@@

\usepackage{amsmath, amsfonts, microtype, xcolor, tikz, graphicx, hyperref}
\usepackage[ruled, linesnumbered]{algorithm2e}
\usepackage[]{neurips_2019}

@@

If using regularized least-squares, arguably the most commonly employed linear regression technique \citep{hoerl1959optimum, rifkin2007notes}, the solution of B2BR has a closed form solution:
\begin{align}
    \hat G &= (Y^\top Y + \Lambda_Y)^{-1} Y^\top X,\label{eq:solG}\\
    \hat H &=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G,\label{eq:solH}
\end{align}
%
where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices of regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{efron1992bootstrap}.
%
After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.

@@

Recall the rank factorisation property of the pseudo-inverse \citep{ben2003generalize}: if $A=BC$ with $A\in\mathbb{R}^{m\times n}$ of rank $k$, $B\in\mathbb{R}^{m\times k}$ and $C\in\mathbb{R}^{k\times n}$, then $A^\dagger=C^\dagger B^\dagger$ and, since $B$ is full rank, $B^\dagger=(B^\top B)^{-1} B^\top$.
%
Using this property on the minimal norm solution $\hat G$ yields
\begin{equation}
\begin{aligned}
(XEF)^\dagger X&=((XK^\top)(KF))^\dagger X\\
&=(KF)^\dagger ((XK^\top)^\top (XK^\top))^{-1} (XK^\top)^\top X\\
&=(KF)^\dagger (KX^\top XK^\top)^{-1} KX^\top X
\end{aligned}
\end{equation}

@@

In regression analysis, the traditional approach to this problem employs a $t$-test to check whether the regression coefficients differ from zero \citep{student1908probable}.
%
However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
%
Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
%
More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{sonquist_morgan} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{E}$, set to one all the diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the diagonal entries.

@@

Back-to-back regression is closely related to Canonical Correlation Analysis \citep{cca_hotelling}. 
%
While CCA computes the eigenvector decomposition of \eqref{eq:solH}, while we compute its diagonal.
%
If \eqref{eq:solH} is diagonal, as in our model \eqref{eq:model}, the variables in $X$ are the eigenvectors.
%
If all the associated eigenvalues are either $0$ or $1$, then all the variables are either part of the $Y$ subspace, or orthogonal to it, respectively.
%
This is because these eigenvalues correspond to the squared cosine of the angle between the corresponding variable in $X$ and the $Y$ subspace.
%
Thus, back-to-back regression is equivalent to CCA under the constraint that the subspaces spanned by $X$ and $Y$ are perpendicular, $X$ is spanned by vectors of $Y$ and vectors orthogonal to $Y$, and that the parallel and orthogonal components of $X$ are spanned by disjoint sets of features of $X$.

@@

\bibliographystyle{plain}
\bibliography{paper}

### 2019-05-23 paper/synthetic.tex

We compare the performance of B2BR against five competing methods:
%
Ridge regression [CITE], multitask Lasso [CITE], Partial Least Squares or PLS [CITE], Canonical Correlation Analysis or CCA [CITE], and Reduced Rank Regression or RRR [CITE].
%
Each of these methods estimates a matrix of coefficients $\hat{W} \in \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$ using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
%
For B2BR, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
%
We employ $5$-fold cross-validation to select the optimal hyper-parameters of each method.

>>>>

We compare the performance of B2BR against five competing methods:
%
Ridge regression \citep{hoerl1959optimum}, multitask Lasso \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls, tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and Reduced Rank Regression or RRR \citep{Izenman_rrr}.
%
Each of these methods estimates a matrix of coefficients $\hat{W} \in \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$ using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
%
For B2BR, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
%
We employ $5$-fold cross-validation to select the optimal hyper-parameters of each method.
